formula1

### 2020-06-21 inst/f1points.Rmd
restructuring

---
documentclass: jss
author:
  - name: Robin K. S. Hankin
    affiliation: Auckland University of Technology
    address: >
      2-14 Wakefield Street
      Auckland, New Zealand
    email: \email{hankin.robin@gmail.com}
    url: http://aut.ac.nz/
title:
  formatted: "An Objective Assessment of Points Scorings in Formula 1 Motor Racing"
  plain:     "An Objective Assessment of Points Scorings in Formula 1 Motor Racing"
  # For running headers, if needed
  short:     "Formula 1 Points Scoring"
abstract: >
  The points scoring system of Formula 1 motor racing is a long-standing
  contention among fans and competitors, with changes to the points
  scored occurring frequently.  Here, I apply generalized Plackett-Luce
  likelihood functions to Formula 1 racing, and assess historical points
  scoring systems objectively.  In a well-defined sense, the points
  scoring system of 1985 is the best of the historical ones.  I make
  similar observations about related competitive situations including
  the Eurovision Song contest and XXX.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

@@

Formula 1 racing is an important and prestigious motor sport
\citep{moul2009}.  Season ranking is based on a points system; points
are awarded to drivers and teams based on race order, and accumulate
additively.  The current points system, for example, is
(25,18,15,12,10,8,6,4,2,1) [that is, the winner is awarded 25 points,
second place 18 points, and so on], although since 2019 the fastest
lap attracts one point.

@@

The points system of Formula 1 Motor racing is a contentious topic;
there have been many changes since systems has changed many times,
although some changes involved only bonus points (typically, one point
for fastest lap), which are not considered here.  The inaugural
system, used from 1950 to 1959 was $(8,6,4,3,2)$, thus crediting only
the first five finishers, and the current system of
$(25,18,15,12,10,8,6,4,2,1)$ credits the first 10.  Arguably these two
systems could introduce different rational behaviour under zero-sum
assumptions: if, in a race, a driver knows he will place seventh under
a low-risk strategy but may place sixth by dint of driving more
aggressively, the low-risk strategy might be rational under the first
points system (which does not reward the extra ranking), but not under
the second, which does.  Still, it is reasonable to assume that each
driver strives to maximise his rank.

@@

```{r,echo=FALSE,print=FALSE}
top <- 11
```

@@

@@

\begin{itemize}
\item Exponential decay $(1,r,r^2,r^3,\ldots)$ for some fixed $r<1$
\item Equal points to the top $n$ competitors
$(\underbrace{1,1,\ldots ,1}_{n},0,0,\ldots)$, noting that
$n=1$ corresponds to a ``winner takes all''
\item Linear decrease $(r,r-1,r-2,\ldots,3,2,1,0)$ for integer $0<r<n$
\item Zipf's law $(1,1/2,1/3,1/4,\ldots)$
\item The current Formula 1 system $(25,18,15,12,10,8,6,4,2,1)$
\end{itemize}

Applying Zipf's law, for example, to the 2017 table we would rank the
drivers as follows:

@@

compare the real:

Thus these two systems agree on the first three places but fourth is
awarded to Ricciardo and Raikkonen respectively.

@@

We now assess the various points systems discussed above, using 1000
_in silico_ trials:

@@

however,

`r sin(1)`

>>>>

---
documentclass: jss
author:
  - name: Robin K. S. Hankin
    affiliation: Auckland University of Technology
    address: >
      2-14 Wakefield Street
      Auckland, New Zealand
    email: \email{hankin.robin@gmail.com}
    url: http://aut.ac.nz/
title:
  formatted: "An Objective Assessment of Points Scorings in Formula 1 Motor Racing"
  plain:     "An Objective Assessment of Points Scorings in Formula 1 Motor Racing"
  # For running headers, if needed
  short:     "Formula 1 Points Scoring"
abstract: >
  The point scoring system of Formula 1 motor racing is a long-standing
  contention among fans and competitors, with changes to the points
  scored occurring frequently.  Here, I apply generalized Plackett-Luce
  likelihood functions to Formula 1 racing, and assess historical points
  scoring systems objectively.  In a well-defined sense, the points
  scoring system of 1985 is the best of the historical ones.  I make
  similar observations about related competitive situations including
  the Eurovision Song contest and XXX.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

@@

Formula 1 racing is an important and prestigious motor sport
\citep{moul2009}.  Season ranking is based on a points allocation
system wherein competitors are awarded points based on race finishing
order; points accumulate additively.  The overall competition winner
is the competitor who accumulates the most points after the final
race.

However, in the case of Formula 1 Motor Racing, the points system is
the subject of much controversy, having changed often since the
competition's inauguration in 1950 when the points allocation, used
from 1950 to 1959 was $(8,6,4,3,2)$, thus crediting only the first
five finishers.  As of 2020, the current system of
$(25,18,15,12,10,8,6,4,2,1)$ credits the first 10.  Arguably these two
systems could introduce different rational behaviour under zero-sum
assumptions: if, in a race, a driver knows he will place seventh under
a low-risk strategy but may place sixth by dint of driving more
aggressively, the low-risk strategy might be rational under the first
points system (which does not reward the extra ranking), but not under
the second, which does.  Still, it is reasonable to assume that each
driver strives to maximise his rank.

@@

@@

```{r,echo=FALSE,print=FALSE}
top <- 11
points_inaugural <- c(8,6,4,3,2)
```

@@

## Points systems

@@

\begin{itemize}
\item Equal points to the top $n$ competitors
$(\underbrace{1,1,\ldots ,1}_{r},0,0,\ldots)$, noting that
$n=r$ corresponds to a ``winner takes all'' system
\item Linear decrease $(r,r-1,r-2,\ldots,3,2,1,0)$ for integer $0<r<n$
\item Exponential decay $(1,r,r^2,r^3,\ldots)$ for some fixed $r<1$
\item Zipf's law $(1,1/2,1/3,1/4,\ldots)$
\item The current Formula 1 system $(25,18,15,12,10,8,6,4,2,1)$
\item The inaugural Formula 1 system $(8,6,4,3,2)$
\end{itemize}

Applying the current points system, for example, to the 2017 table we
would rank the drivers as follows:

(this happens to be identical to the ranking after the extra point was
awarded for fastest lap).  However, if we were to use Zipf's law, the
ranking would be:

@@

Thus these two systems agree on the first three places but fourth is
awarded to Raikkonen under the F1 system and Ricciardo under Zipf.

Now compare the likelihood ranking:

```{r,label=likelihoodrankingshower,echo=FALSE}
jj <- mH
jj <- sort(jj,decreasing=TRUE)
jj[] <- seq_along(jj)
jj
```

@@

We now assess the various points systems discussed above, using 1000
_in silico_ trials.

@@

# Discussion

Points systems are a source of lively debate from many perspectives.
The point of a point allocation system is to rank competitors in an
objective way; but

Here I treat total points scored as a random variable and compare
different point allocation schemes against both one another and

but until now there is no objective way of comparing points systems.
if one considers order tables to be a random variable.  Because the
numerical value of points awarded for first, second, third etc, does
not affect driver behaviour it is a reasonable requirement that

Here, I assess
several plausible points systems including the one actually used in
Formula 1.  By several criteria, the
